---
title: "CiteFuse: getting start"
output: 
  rmarkdown::html_vignette:
    number_sections: yes
vignette: >
  %\VignetteIndexEntry{scClassify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# Introduction 

`CiteFuse` is a computational framework that implements a suite of methods and tools for CITE-seq data from pre-processing through to integrative analytics. This includes doublet detection, network-based modality integration, cell type clustering, differential RNA and ADT expression analysis, ADT evaluation, ligand-receptor interaction analysis, and interactive web-based visualisation of the analyses. this vignette demostrates the usage of `CiteFuse` using a small subset data of CITE-seq data from PBMC as an example (Mimitou et al., 2019).


```{r}
library(CiteFuse)
library(scater)
library(SingleCellExperiment)
library(DT)
```


```{r}
load("data/CITEseq_example.rda", verbose = T)
```

Here, we started from three matrices of RNA, ADT and HTO, which have common cell names. The `preprocessing` function will utilise the three matrices and its common cell names to create a `SingleCellExperiment` object, which stores RNA data in `assay`, and `ADT`, `HTO` data in `altExp` slot.

```{r}
sce_citeseq <- preprocessing(list(RNA = rna_subset,
                                  ADT = adt_subset,
                                  HTO = hto_subset))

sce_citeseq
```


# Detecting both cross- and within-sample doublets using `CiteFuse`



## HTO Normalisation and Visualisation


The function `notmaliseExprs` is used to scale the alternative expression. Here, we used it to perform log-transformation for the `HTO` expression, by setting `transform = "log"`.


```{r}
sce_citeseq <- normaliseExprs(sce = sce_citeseq, 
                              altExp_name = "HTO", 
                              transform = "log")
```


Then we can perform dimension reduction on the `HTO` expression using `runTSNE` or `runUMAP`, then use `visualiseDim` function to visualise the reduced dimension plot.


```{r fig.height=6, fig.width=6}

sce_citeseq <- scater::runTSNE(sce_citeseq, 
                               altexp = "HTO", 
                               name = "TSNE_HTO", 
                               pca = TRUE)


visualiseDim(sce_citeseq,
             dimNames = "TSNE_HTO") + labs(title = "tSNE (HTO)")

sce_citeseq <- scater::runUMAP(sce_citeseq, 
                               altexp = "HTO", 
                               name = "UMAP_HTO")


visualiseDim(sce_citeseq,
             dimNames = "UMAP_HTO") + labs(title = "UMAP (HTO)")
```




## Doublet Identification (cross sample)

We first identify the cross-sample doublets via function `crossSampleDoublets`.


```{r}
sce_citeseq <- crossSampleDoublets(sce_citeseq)
```


The results of the cross sample doublets are then saved in the `colData` as `doubletClassify_between_label` and `doubletClassify_between_class`.

```{r}
table(sce_citeseq$doubletClassify_between_label)
table(sce_citeseq$doubletClassify_between_class)
```

We can then highlights the cross-sample doublets in our tSNE plot of HTO expression.

```{r fig.height=6, fig.width=6}
visualiseDim(sce_citeseq, 
             dimNames = "TSNE_HTO", 
             colour_by = "doubletClassify_between_label")
```


Further, `plotHTO` function allows us to plot the pairwise scatter HTO expression.

```{r fig.height=12, fig.width=8}
plotHTO(sce_citeseq, 1:4)
```


## Doublet Identification (within samples)

We then identify the within-sample doublets via function `withinSampleDoublets`.


```{r}
sce_citeseq <- withinSampleDoublets(sce_citeseq,
                                    minPts = 10)
```

The results of the cross sample doublets are then saved in the `colData` as `doubletClassify_within_label` and `doubletClassify_within_class`.


```{r}
table(sce_citeseq$doubletClassify_within_label)
table(sce_citeseq$doubletClassify_within_class)
```



```{r fig.height=6, fig.width=6}
visualiseDim(sce_citeseq, 
             dimNames = "TSNE_HTO", 
             colour_by = "doubletClassify_within_label")
```



Finally, we can filter out the doublets cells (Both within and between batches) for the downstream analysis.


```{r}
sce_citeseq <- sce_citeseq[, sce_citeseq$doubletClassify_within_class == "Singlet" & sce_citeseq$doubletClassify_between_class == "Singlet"]
sce_citeseq
```


# Clustering



## Performing SNF

```{r}
sce_citeseq <- scater::normalize(sce_citeseq)
system.time(sce_citeseq <- CiteFuse(sce_citeseq))
```


## Performing spectral clustering

```{r}
SNF_W_clust <- spectralClustering(metadata(sce_citeseq)[["SNF_W"]], K = 20)
plot(SNF_W_clust$eigen_values)
which.max(abs(diff(SNF_W_clust$eigen_values)))

SNF_W_clust <- spectralClustering(metadata(sce_citeseq)[["SNF_W"]], K = 5)
sce_citeseq$SNF_W_clust <- as.factor(SNF_W_clust$labels)



SNF_W1_clust <- spectralClustering(metadata(sce_citeseq)[["ADT_W"]], K = 5)
sce_citeseq$ADT_clust <- as.factor(SNF_W1_clust$labels)

SNF_W2_clust <- spectralClustering(metadata(sce_citeseq)[["RNA_W"]], K = 5)
sce_citeseq$RNA_clust <- as.factor(SNF_W2_clust$labels)
```


## Visualisation

```{r fig.height=12, fig.width=8}

sce_citeseq <- reducedDimSNF(sce_citeseq, method = "tSNE", dimNames = "tSNE_joint")

g1 <- visualiseDim(sce_citeseq, dimNames = "tSNE_joint", colour_by = "SNF_W_clust") +
  labs(title = "tSNE (SNF clustering)")
g2 <- visualiseDim(sce_citeseq, dimNames = "tSNE_joint",  colour_by = "ADT_clust") +
  labs(title = "tSNE (ADT clustering)")
g3 <- visualiseDim(sce_citeseq, dimNames = "tSNE_joint",  colour_by = "RNA_clust") +
  labs(title = "tSNE (RNA clustering)")

library(gridExtra)
grid.arrange(g3, g2, g1, ncol = 2)
```


## Louvain Clustering

```{r fig.height=8, fig.width=8}
SNF_W_louvain <- igraphClustering(sce_citeseq, method = "louvain")
table(SNF_W_louvain)

sce_citeseq$SNF_W_louvain <- as.factor(SNF_W_louvain)

visualiseDim(sce_citeseq, dimNames = "tSNE_joint", colour_by = "SNF_W_louvain") +
  labs(title = "tSNE (SNF louvain clustering)")
```


```{r fig.height = 6, fig.width = 6}
visualiseKNN(sce_citeseq, colour_by = "SNF_W_louvain")
```


# Differential Expression Analysis

## Exploration of features expression 

```{r fig.height = 8, fig.width = 8}
g1 <- visualiseExprs(sce_citeseq, 
                     plot = "boxplot", 
                     feature_subset = c("hg19_CD2", "hg19_CD4", "hg19_CD8A", "hg19_CD19"))
g2 <- visualiseExprs(sce_citeseq, 
                     plot = "violin", 
                     feature_subset = c("hg19_CD2", "hg19_CD4", "hg19_CD8A", "hg19_CD19"))
g3 <- visualiseExprs(sce_citeseq, 
                     plot = "jitter", 
                     feature_subset = c("hg19_CD2", "hg19_CD4", "hg19_CD8A", "hg19_CD19"))
g4 <- visualiseExprs(sce_citeseq, 
                     plot = "density", 
                     feature_subset = c("hg19_CD2", "hg19_CD4", "hg19_CD8A", "hg19_CD19"))

library(gridExtra)
grid.arrange(g1, g2, g3, g4, ncol = 2)
```


```{r fig.height = 8, fig.width = 8}
sce_citeseq <- normaliseExprs(sce_citeseq, altExp_name = "ADT", transform = "log")
visualiseExprs(sce_citeseq, 
               altExp_name = "ADT", 
               n = 30)
g1 <- visualiseExprs(sce_citeseq, 
                     altExp_name = "ADT", 
                     plot = "violin", n = 5)
g2 <- visualiseExprs(sce_citeseq, altExp_name = "ADT", 
                     plot = "jitter", 
                     feature_subset = c("CD2", "CD8", "CD4", "CD19"))
g3 <- visualiseExprs(sce_citeseq, altExp_name = "ADT", 
                     plot = "density", 
                     feature_subset = c("CD2", "CD8", "CD4", "CD19"))

grid.arrange(g1, g2, g3, ncol = 2)
```


```{r fig.height = 8, fig.width = 8}
visualiseExprs(sce_citeseq, altExp_name = "ADT", 
               plot = "boxplot", 
               feature_subset = c("CD2", "CD8", "CD4", "CD19")) + 
  facet_wrap(~sce_citeseq$SNF_W_louvain, nrow = 7)
```


```{r fig.height = 8, fig.width = 8}
visualiseExprs(sce_citeseq, altExp_name = "ADT", 
               plot = "pairwise", 
               feature_subset = c("CD4", "CD8"))

visualiseExprs(sce_citeseq, altExp_name = "ADT", 
               plot = "pairwise", 
               feature_subset = c("CD45RA", "CD4", "CD8"), threshold = rep(4, 3))
```


## Perform DE analysis using Wilcoxon Rank Sum test

```{r}
rna_DEgenes_all <- DEgenes(sce_citeseq,
                           group = sce_citeseq$SNF_W_louvain,
                           return_all = TRUE,
                           exprs_pct = 0.5)


rna_DEgenes <- selectDEgenes(rna_DEgenes_all)
datatable(format(do.call(rbind, rna_DEgenes), digits = 2))
```


```{r}
adt_DEgenes_all <- DEgenes(sce_citeseq,
                           altExp_name = "ADT",
                           group = sce_citeseq$SNF_W_louvain,
                           exprs_threshold = 4,
                           return_all = TRUE,
                           exprs_pct = 0.1)

adt_DEgenes <- selectDEgenes(adt_DEgenes_all)



datatable(format(do.call(rbind, adt_DEgenes), digits = 2))
```


## Visualising DE results

### circlepackPlot

```{r fig.height = 10, fig.width = 10}
rna_DEgenes <- lapply(rna_DEgenes, function(x){
  x$name <- gsub("hg19_", "", x$name)
  x
})
DEbubblePlot(list(RNA = rna_DEgenes, ADT = adt_DEgenes))
```


### DEcomparisonPlot

```{r fig.height = 10, fig.width = 10}
rna_list <- c("hg19_CD4",
              "hg19_CD8A",
              "hg19_HLA-DRB1",
              "hg19_ITGAX",
              "hg19_NCAM1",
              "hg19_CD27",
              "hg19_CD19")

adt_list <- c("CD4", "CD8", "MHCII (HLA-DR)", "CD11c", "CD56", "CD27", "CD19")

feature_list <- list(RNA = rna_list, ADT = adt_list)
de_list <- list(RNA = rna_DEgenes_all, ADT = adt_DEgenes_all)

DEcomparisonPlot(de_list = de_list,
                 feature_list = feature_list)
```


# ADT importance evaluation

```{r fig.height = 8, fig.width = 8}
sce_citeseq <- importanceADT(sce_citeseq, 
                             group = sce_citeseq$SNF_W_louvain,
                             subsample = T)

visImportance(sce_citeseq, plot = "boxplot")
visImportance(sce_citeseq, plot = "heatmap")

sort(metadata(sce_citeseq)[["importanceADT"]], decreasing = TRUE)[1:20]
```


Select the ADT with scores greater than 5 and perform SNF again

```{r}
subset_adt <- names(which(metadata(sce_citeseq)[["importanceADT"]] > 5))
subset_adt

system.time(sce_citeseq <- CiteFuse(sce_citeseq,
                                    ADT_subset = subset_adt,
                                    metadata_names = c("W_SNF_adtSubset1",
                                                       "W_ADT_adtSubset1",
                                                       "W_RNA"
                                    )))

SNF_W_clust_adtSubset1 <- spectralClustering(metadata(sce_citeseq)[["W_SNF_adtSubset1"]], K = 5)
sce_citeseq$SNF_W_clust_adtSubset1 <- as.factor(SNF_W_clust_adtSubset1$labels)


library(mclust)
adjustedRandIndex(sce_citeseq$SNF_W_clust_adtSubset1, sce_citeseq$SNF_W_clust)
```


# Gene - ADT network

```{r fig.height = 8, fig.width = 8}
RNA_feature_subset <- unique(as.character(unlist(lapply(rna_DEgenes_all, "[[", "name"))))
ADT_feature_subset <- unique(as.character(unlist(lapply(adt_DEgenes_all, "[[", "name"))))

geneADTnetwork(sce_citeseq,
               RNA_feature_subset = RNA_feature_subset,
               ADT_feature_subset = ADT_feature_subset,
               cor_method = "pearson",
               network_layout = igraph::layout_with_fr)
```


# RNA Ligand - ADT Receptor Analysis


```{r}
load("data/lr_pairs_example.rda", verbose = T)
head(lr_pair_subset)

sce_citeseq <- normaliseExprs(sce = sce_citeseq, 
                              altExp_name = "ADT", 
                              transform = "zi_minMax")

sce_citeseq <- normaliseExprs(sce = sce_citeseq, 
                              altExp_name = "none", 
                              exprs_value = "logcounts",
                              transform = "minMax")

sce_citeseq <- ligandReceptorTest(sce = sce_citeseq,
                                  ligandReceptor_list = lr_pair_subset,
                                  cluster = sce_citeseq$SNF_W_louvain,
                                  RNA_exprs_value = "minMax",
                                  altExp_name = "ADT",
                                  altExp_exprs_value = "zi_minMax",
                                  num_permute = 1000,
                                  ncores = 1) 
```


```{r fig.height=10, fig.width=8}
visLigandReceptor(sce_citeseq, type = "pval_heatmap")

```


```{r fig.height=12, fig.width=6}
visLigandReceptor(sce_citeseq, type = "pval_dotplot")
```


```{r fig.height=8, fig.width=8}
visLigandReceptor(sce_citeseq, type = "group_network")
```


```{r fig.height=8, fig.width=8}
visLigandReceptor(sce_citeseq, type = "group_heatmap")
```


```{r fig.height=8, fig.width=8}
visLigandReceptor(sce_citeseq, type = "lr_network")
```


# SessionInfo

```{r}
sessionInfo()
```





